### 2015-12-21 abstract.tex
Polished grammar and typography.

We design algorithms for online linear optimization that have optimal regret
and at the same time do not need to know any upper or lower bounds on the norm
of the loss vectors.  We achieve adaptiveness to the norms of the loss vectors by scale
invariance, i.e., our algorithms make exactly the same decisions if the
sequence of loss vectors is multiplied by any positive constant.  Our
algorithms work for any decision set, bounded or unbounded.  For unbounded
decisions sets, these are the first adaptive algorithms for online linear
optimization with a non-vacuous bound.

>>>>

We design algorithms for online linear optimization that have optimal regret
and at the same time do not need to know any upper or lower bounds on the norm
of the loss vectors.  We achieve adaptiveness to the norms of the loss vectors
by scale invariance, i.e., our algorithms make exactly the same decisions if
the sequence of loss vectors is multiplied by any positive constant.  One of
our algorithms works for any decision set, bounded or unbounded.  For unbounded
decisions sets, it is the first adaptive algorithm for online linear
optimization with a non-vacuous bound.

### 2015-12-21 ada-ftrl.tex
typos and small changes
Polished grammar and typography.

Assume that $K$ is bounded and that $R(w)$ is a strongly convex lower
semi-continuous function bounded from above.  We instantiate
Algorithm~\ref{algorithm:ftrl-varying-regularizer} with the sequence of
regularizers
\begin{equation}
\label{equation:ada-ftrl}
R_t(w) = \Delta_{t-1} R(w)
\quad \text{where}
\quad \Delta_{t}=\sum_{i=1}^{t} \Delta_{i-1} \Breg_{R^*}\left(- \frac{L_i}{\Delta_{i-1}}, -\frac{L_{i-1}}{\Delta_{i-1}}\right) \; .
\end{equation}

@@

\begin{theorem}[Regret Bound]
\label{theorem:ada-ftrl-regret-bound}
Suppose $K \subseteq V$ is a non-empty bounded closed convex subset. Let $D =
\sup_{x,y \in K} \|x - y\|$ be its diameter with respect to a norm $\|\cdot\|$.
Suppose that the regularizer $R:K \to \R$ is a non-negative lower
semi-continuous function that is $\lambda$-strongly convex with respect to
$\|\cdot\|$ and is bounded from above.  The regret of \textsc{AdaFTRL} satisfies
$$
\Regret_T(u)
\le \sqrt{3} \max\left\{D, \frac{1}{\sqrt{2\lambda}} \right\}
    \sqrt{\sum_{t=1}^T \|\ell_t\|_*^2} \left(1 + R(u) \right) \; .
$$
\end{theorem}

@@

\begin{corollary}[Regret Bound]
\label{corollary:ada-ftrl-regret-bound}
Suppose $K \subseteq V$ is a non-empty bounded closed convex subset. Suppose
$f:K \to \R$ is a non-negative lower semi-continuous function that is
$1$-strongly convex with respect to $\|\cdot\|$ and is bounded from above.  The
regret of \textsc{AdaFTRL} with regularizer
$$
R(w) = \frac{f(w)}{16 \cdot \sup_{v \in K} f(v)}
\qquad \text{satisfies} \qquad
\Regret_T \le
5.3 \sqrt{\sup_{v \in K} f(v) \sum_{t=1}^T \|\ell_t\|_*^2} \; .
$$
\end{corollary}

@@

\begin{lemma}[Recurrence]
\label{lemma:recurrence-solution}
Let $\{a_t\}_{t=1}^\infty$ be any sequence of non-negative real numbers.
Suppose that $\{\Delta_t\}_{t=0}^\infty$ is a sequence of non-negative real
numbers satisfying
$$
\Delta_0 = 0 \qquad \text{and} \qquad
\Delta_t \le \Delta_{t-1} + \min \left\{ a_t, \ \frac{a_t^2}{\Delta_{t-1}} \right\} \quad \text{for any $t \ge 1$} \; .
$$
Then, for any $T \ge 0$,
$$
\Delta_T \le \sqrt{3 \sum_{t=1}^T a_t^2} \; .
$$
\end{lemma}
%
Proof of the Lemma~\ref{lemma:recurrence-solution} is deferred
to~\ref{section:ada-ftrl-proofs}.  Theorem~\ref{theorem:ada-ftrl-regret-bound}
follows from
Lemmas~\ref{lemma:initial-regret-bound},~\ref{lemma:gap-recurrence}~and~\ref{lemma:recurrence-solution}.

>>>>

Assume that $K$ is bounded and that $R:K \to \R$ is a strongly convex lower
semi-continuous function bounded from above.  We instantiate
Algorithm~\ref{algorithm:ftrl-varying-regularizer} with the sequence of
regularizers
\begin{equation}
\label{equation:ada-ftrl}
R_t(w) = \Delta_{t-1} R(w)
\quad \text{where}
\quad \Delta_{t}=\sum_{i=1}^{t} \Delta_{i-1} \Breg_{R^*}\left(- \frac{L_i}{\Delta_{i-1}}, -\frac{L_{i-1}}{\Delta_{i-1}}\right) \; .
\end{equation}

@@

\begin{theorem}[Regret Bound]
\label{theorem:ada-ftrl-regret-bound}
Suppose $K \subseteq V$ is a non-empty bounded closed convex set. Let $D =
\sup_{x,y \in K} \|x - y\|$ be its diameter with respect to a norm $\|\cdot\|$.
Suppose that the regularizer $R:K \to \R$ is a non-negative lower
semi-continuous function that is $\lambda$-strongly convex with respect to
$\|\cdot\|$ and is bounded from above.  The regret of \textsc{AdaFTRL}
satisfies
$$
\Regret_T(u)
\le \sqrt{3} \max\left\{D, \frac{1}{\sqrt{2\lambda}} \right\}
    \sqrt{\sum_{t=1}^T \|\ell_t\|_*^2} \left(1 + R(u) \right) \; .
$$
\end{theorem}

@@

\begin{corollary}[Regret Bound]
\label{corollary:ada-ftrl-regret-bound}
Suppose $K \subseteq V$ is a non-empty bounded closed convex set. Suppose $f:K
\to \R$ is a non-negative lower semi-continuous function that is $1$-strongly
convex with respect to $\|\cdot\|$ and is bounded from above.  The regret of
\textsc{AdaFTRL} with regularizer
$$
R(w) = \frac{f(w)}{16 \cdot \sup_{v \in K} f(v)}
\qquad \text{satisfies} \qquad
\Regret_T \le
5.3 \sqrt{\sup_{v \in K} f(v) \sum_{t=1}^T \|\ell_t\|_*^2} \; .
$$
\end{corollary}

@@

\begin{lemma}[Solution of the recurrence]
\label{lemma:recurrence-solution}
Let $\{a_t\}_{t=1}^\infty$ be any sequence of non-negative real numbers.
Suppose that $\{\Delta_t\}_{t=0}^\infty$ is a sequence of non-negative real
numbers satisfying
$$
\Delta_0 = 0 \qquad \text{and} \qquad
\Delta_t \le \Delta_{t-1} + \min \left\{ a_t, \ \frac{a_t^2}{\Delta_{t-1}} \right\} \quad \text{for any $t \ge 1$} \; .
$$
Then, for any $T \ge 0$,
$$
\Delta_T \le \sqrt{3 \sum_{t=1}^T a_t^2} \; .
$$
\end{lemma}
%
Proof of the Lemma~\ref{lemma:recurrence-solution} is deferred
to~\ref{section:ada-ftrl-proofs}.  Theorem~\ref{theorem:ada-ftrl-regret-bound}
follows from
Lemmas~\ref{lemma:initial-regret-bound},~\ref{lemma:gap-recurrence}~and~\ref{lemma:recurrence-solution}.

### 2015-12-21 introduction.tex
Polished grammar and typography.

There is a clear practical need to design algorithms that adapt automatically
to norms of the loss vectors.  A natural, yet overlooked, design method to
achieve this type of adaptivity is by insisting to have a \textbf{scale-free}
algorithm.  That is, the sequence of decisions of the algorithm does not change
if the sequence of loss vectors is multiplied by a positive constant.

The most important property of scale-free algorithms is that both their loss
and their regret is scales linearly with the maximum norm of the loss vector
appearing in the sequence.

@@

Majority of existing algorithms for OLO are based on two generic algorithms:
\textsc{Follow The Regularizer Leader} (\textsc{FTRL}) and \textsc{Mirror
Descent} (\textsc{MD}). \textsc{FTRL} dates back to the potential-based
forecaster in \cite[Chapter~11]{Cesa-Bianchi-Lugosi-2006} and its theory was
developed in \cite{Shalev-Shwartz-2007}.  The name \textsc{Follow The
Regularized Leader} comes from~\cite{Abernethy-Hazan-Rakhlin-2008}.
Independently, the same algorithm was proposed in~\cite{Nestorov-2009} for
convex optimization under the name \textsc{Dual Averaging} and later
rediscovered in~\cite{Xiao-2010} for online convex optimization. The general
version with time-varying regularizers was analyzed in
\cite{Orabona-Crammer-Cesa-Bianchi-2014}. \textsc{MD} was originally proposed
in \cite{Nemirovski-Yudin-1983} and later analyzed in~\cite{Beck-Teboulle-2003}
for convex optimization. In the online learning literature it makes the first
appearance, under a different name, in \cite{Kivinen-Warmuth-1997}. Good
description of \textsc{MD} is given in~\cite{Rakhlin-Sridharan-2009}.

>>>>

There is a clear practical need to design algorithms that adapt automatically
to norms of the loss vectors.  A natural, yet overlooked, design method to
achieve this type of adaptivity is by insisting to have a \textbf{scale-free}
algorithm.  That is, the sequence of decisions of the algorithm does not change
if the sequence of loss vectors is multiplied by a positive constant.  The most
important property of scale-free algorithms is that both their loss and their
regret scale linearly with the maximum norm of the loss vector appearing in the
sequence.

@@

Majority of existing algorithms for OLO are based on two generic algorithms:
\textsc{Follow The Regularizer Leader} (\textsc{FTRL}) and \textsc{Mirror
Descent} (\textsc{MD}). \textsc{FTRL} dates back to the potential-based
forecaster in \cite[Chapter~11]{Cesa-Bianchi-Lugosi-2006} and its theory was
developed in \cite{Shalev-Shwartz-2007}.  The name \textsc{Follow The
Regularized Leader} comes from~\cite{Abernethy-Hazan-Rakhlin-2008}.
Independently, the same algorithm was proposed in~\cite{Nestorov-2009} for
convex optimization under the name \textsc{Dual Averaging} and later
rediscovered in~\cite{Xiao-2010} for online convex optimization. The general
version with time-varying regularizers was analyzed in
\cite{Orabona-Crammer-Cesa-Bianchi-2014}. \textsc{MD} was originally proposed
in \cite{Nemirovski-Yudin-1983} and later analyzed in~\cite{Beck-Teboulle-2003}
for convex optimization. In the online learning literature it makes the first
appearance, under a different name, in \cite{Kivinen-Warmuth-1997}. A good
description of \textsc{MD} is given in~\cite{Rakhlin-Sridharan-2009}.

### 2015-12-21 limits.tex
Polished grammar and typography.

\begin{proof}
Using property 3 of Proposition~\ref{proposition:conjugate-properties} we can write
the divergence
\begin{align*}
a \Breg_{R^*}(x/a, y/a) & = a R^*(x/a) - a R^*(y/a) - \langle x - y, \grad R^*(y/a) \rangle \\
& =
 a \left[ \langle x/a, \grad R^*(x/a) \rangle - R(\grad R^*(x/a)) \right] \\
& \qquad - a \left[ \langle y/a, \grad R^*(y/a) \rangle - R(\grad R^*(y/a)) \right]
- \langle x - y, \grad R^*(y/a) \rangle \\
& =
\langle x, \grad R^*(x/a) - \grad R^*(y/a) \rangle - a R(\grad R^*(x/a))
+ a R(\grad R^*(y/a)) \; .
\end{align*}
Property 2 of Proposition~\ref{proposition:conjugate-properties} implies that
\begin{align*}
u = \lim_{a \to 0^+} \grad R^*(x/a) & = \lim_{a \to 0^+} \argmin_{w \in K} \left( a R(w) - \langle x, w \rangle \right) \; , \\
v = \lim_{a \to 0^+} \grad R^*(y/a) & = \lim_{a \to 0^+} \argmin_{w \in K} \left( a R(w) - \langle y, w \rangle \right) \; .
\end{align*}
The limits on the right exist because of compactness of $K$. They are simply
the minimizers $u= \argmin_{w \in K} - \langle x, w \rangle$ and $v= \argmin_{w
\in K} - \langle y, w \rangle$ where ties in $\argmin$ are broken according to
smaller value of $R(w)$.

>>>>

\begin{proof}
Using property 3 of Proposition~\ref{proposition:conjugate-properties} we can write
the divergence
\begin{align*}
a \Breg_{R^*}(x/a, y/a) & = a R^*(x/a) - a R^*(y/a) - \langle x - y, \grad R^*(y/a) \rangle \\
& =
 a \left[ \langle x/a, \grad R^*(x/a) \rangle - R(\grad R^*(x/a)) \right] \\
& \qquad - a \left[ \langle y/a, \grad R^*(y/a) \rangle - R(\grad R^*(y/a)) \right]
- \langle x - y, \grad R^*(y/a) \rangle \\
& =
\langle x, \grad R^*(x/a) - \grad R^*(y/a) \rangle - a R(\grad R^*(x/a))
+ a R(\grad R^*(y/a)) \; .
\end{align*}
Property 2 of Proposition~\ref{proposition:conjugate-properties} implies that
\begin{align*}
u = \lim_{a \to 0^+} \grad R^*(x/a) & = \lim_{a \to 0^+} \argmin_{w \in K} \left( a R(w) - \langle x, w \rangle \right) \; , \\
v = \lim_{a \to 0^+} \grad R^*(y/a) & = \lim_{a \to 0^+} \argmin_{w \in K} \left( a R(w) - \langle y, w \rangle \right) \; .
\end{align*}
The limits on the right exist according to
Lemma~\ref{lemma:prediction-limit-existence}.  They are simply the minimizers
$u = \argmin_{w \in K} - \langle x, w \rangle$ and $v= \argmin_{w \in K} -
\langle y, w \rangle$ where ties in $\argmin$ are broken according to smaller
value of $R(w)$.

### 2015-12-21 mirror-descent.tex
Polished grammar and typography.

The most prominent instance of \textsc{MD} is \textsc{Projected Gradient
Descent} defined by the recurrence
$$
w_{t+1} = \Pi_K(w_t - \eta_t \ell_t)
$$
where $\Pi_K(x) = \argmin_{y \in K} \|x-y\|_2$ is the Euclidean projection to
$K$ and $\eta_t > 0$ is the step size.  \textsc{Projected Gradient Descent} is
an instance of \textsc{MD} with regularizer $R_t(w) =
\frac{1}{2\eta_t}\|w\|_2^2$.  A special case of \textsc{Projected Gradient
Descent} with step size $\eta_t = \frac{1}{\sqrt{t}}$ is \textsc{Generalized
Infinitesimal Gradient Ascent} (GIGA)~\cite{Zinkevich-2003}.  Full-matrix and
diagonal versions of \textsc{AdaGrad with Composite MD
Update}~\cite{Duchi-Hazan-Singer-2011} are an instances of \textsc{MD} for any
bounded $K$ with regularizers
$$
R_t(w) = \frac{1}{2} w^\top \left(\sum_{i=1}^t \ell_i \ell_i^\top \right)^{1/2} \!\!\!\! w
\qquad \text{and} \qquad
R_t(w) = \frac{1}{2} \sum_{j=1}^d w_j^2 \sqrt{ \sum_{i=1}^t \ell_{i,j}^2}
$$
respectively. The diagonal version is simply the per-coordinate construction
applied to the one-dimensional version. When $d=1$, both algorithms are
equivalent to
\begin{equation}
\label{equation:ada-grad-regularizer}
R_t(w) = \frac{\|w\|_2^2}{2} \sqrt{\sum_{i=1}^t \|\ell_i\|_2^2 } \; .
\end{equation}
All three sequences of regularizers give a scale-free algorithm. Regularizer
\eqref{equation:ada-grad-regularizer} can be viewed as a scale-free
generalization of GIGA. That is, when $\|\ell_i\|_2
= 1$ for all $i$, \textsc{MD} with regularizer
\eqref{equation:ada-grad-regularizer} coincides with GIGA.

@@

\begin{theorem}[Regret of Scale-Free Mirror Descent]
\label{theorem:regret-scale-free-mirror-descent}
Suppose $K \subseteq V$ is a non-empty closed convex subset. Suppose that $R:K
\to \R$ is a $\lambda$-strongly convex function with respect to a norm
$\|\cdot\|$.  \textsc{Scale-Free MD} with regularizer $R$ satisfies
for any $u \in K$,
$$
\Regret_T(u)
\le
\left( \frac{1}{\lambda} + \sup_{v \in K} \Breg_R(u,v) \right) \sqrt{\sum_{t=1}^T \|\ell_t\|_*^2} \; .
$$
\end{theorem}

@@

\begin{corollary}[Regret of Scale-Free Mirror Descent]
\label{corollary:regret-scale-free-mirror-descent}
Suppose $K \subseteq V$ is a non-empty bounded closed convex subset.  Suppose
that $f:K \to \R$ is a $1$-strongly convex function with respect to a norm
$\|\cdot\|$.  \textsc{Scale-Free MD} with regularizer
$$
R(w) = \frac{f(w)}{\displaystyle \sqrt{\sup_{u,v \in K} \Breg_f(u,v)}}
\quad \text{satisfies} \quad %
\Regret_T \le 2 \sqrt{\sup_{u,v \in K} \Breg_f(u,v) \sum_{t=1}^T \|\ell_t\|_*^2} \; .
$$
\end{corollary}

@@

The first counter-example is stated as
Theorem~\ref{theorem:first-counter-example} below; proof is
in~\ref{section:mirror-descent-proofs}. The decision set is the whole space
$K=V$ and the regularizer is $R(w) = \frac{1}{2}\|w\|_2^2$. Note that $R(w)$ is
$1$-strongly convex with respect to $\|\cdot\|_2$ and the dual norm of
$\|\cdot\|_2$ is $\|\cdot\|_2$. The corresponding Bregman divergence is
$\Breg_R(u,v) = \frac{1}{2}\|u-v\|_2^2$. The counter-example constructs
sequence of unit-norm loss vectors in the one-dimensional subspace spanned by
the first vector of the standard unit basis.  On such sequence, Zinkevich's
\textsc{GIGA} and both versions of \textsc{AdaGrad MD} have predictions
identical to \textsc{Scale-Free MD}. Hence the lower bound applies to all four
algorithms.

>>>>

The most prominent instance of \textsc{MD} is \textsc{Projected Gradient
Descent} defined by the recurrence
$$
w_{t+1} = \Pi_K \left( w_t - \eta_t \ell_t \right)
$$
where $\Pi_K(x) = \argmin_{y \in K} \|x-y\|_2$ is the Euclidean projection to
$K$ and $\eta_t > 0$ is the step size.  \textsc{Projected Gradient Descent} is
an instance of \textsc{MD} with regularizer $R_t(w) =
\frac{1}{2\eta_t}\|w\|_2^2$.  A special case of \textsc{Projected Gradient
Descent} with step size $\eta_t = \frac{1}{\sqrt{t}}$ is \textsc{Generalized
Infinitesimal Gradient Ascent} (GIGA)~\cite{Zinkevich-2003}.  Full-matrix and
diagonal versions of \textsc{AdaGrad with Composite MD
Update}~\cite{Duchi-Hazan-Singer-2011} are an instances of \textsc{MD} for any
bounded $K$ with regularizers
$$
R_t(w) = \frac{1}{2} w^\top \left(\sum_{i=1}^t \ell_i \ell_i^\top \right)^{1/2} \!\!\!\! w
\qquad \text{and} \qquad
R_t(w) = \frac{1}{2} \sum_{j=1}^d w_j^2 \sqrt{ \sum_{i=1}^t \ell_{i,j}^2}
$$
respectively. The diagonal version is simply the per-coordinate construction
applied to the one-dimensional version. When $d=1$, both algorithms are
equivalent to
\begin{equation}
\label{equation:ada-grad-regularizer}
R_t(w) = \frac{\|w\|_2^2}{2} \sqrt{\sum_{i=1}^t \|\ell_i\|_2^2 } \; .
\end{equation}
All three sequences of regularizers give a scale-free algorithm. Regularizer
\eqref{equation:ada-grad-regularizer} can be viewed as a scale-free
generalization of GIGA. That is, when $\|\ell_i\|_2
= 1$ for all $i$, \textsc{MD} with regularizer
\eqref{equation:ada-grad-regularizer} coincides with GIGA.

@@

\begin{theorem}[Regret of Scale-Free Mirror Descent]
\label{theorem:regret-scale-free-mirror-descent}
Suppose $K \subseteq V$ is a non-empty closed convex set. Suppose that $R:K
\to \R$ is a $\lambda$-strongly convex function with respect to a norm
$\|\cdot\|$.  \textsc{Scale-Free MD} with regularizer $R$ satisfies
for any $u \in K$,
$$
\Regret_T(u)
\le
\left( \frac{1}{\lambda} + \sup_{v \in K} \Breg_R(u,v) \right) \sqrt{\sum_{t=1}^T \|\ell_t\|_*^2} \; .
$$
\end{theorem}

@@

\begin{corollary}[Regret of Scale-Free Mirror Descent]
\label{corollary:regret-scale-free-mirror-descent}
Suppose $K \subseteq V$ is a non-empty bounded closed convex set.  Suppose
that $f:K \to \R$ is a $1$-strongly convex function with respect to a norm
$\|\cdot\|$.  \textsc{Scale-Free MD} with regularizer
$$
R(w) = \frac{f(w)}{\displaystyle \sqrt{\sup_{u,v \in K} \Breg_f(u,v)}}
\quad \text{satisfies} \quad %
\Regret_T \le 2 \sqrt{\sup_{u,v \in K} \Breg_f(u,v) \sum_{t=1}^T \|\ell_t\|_*^2} \; .
$$
\end{corollary}

@@

The first counter-example is stated as
Theorem~\ref{theorem:first-counter-example} below; proof is
in~\ref{section:mirror-descent-proofs}. The decision set is the whole space
$K=V$ and the regularizer is $R(w) = \frac{1}{2}\|w\|_2^2$. Note that $R(w)$ is
$1$-strongly convex with respect to $\|\cdot\|_2$ and the dual norm of
$\|\cdot\|_2$ is $\|\cdot\|_2$. The corresponding Bregman divergence is
$\Breg_R(u,v) = \frac{1}{2}\|u-v\|_2^2$. The counter-example constructs
sequence of unit-norm loss vectors in the one-dimensional subspace spanned by
the first vector of the standard orthnormal basis.  On such sequence,
Zinkevich's \textsc{GIGA} and both versions of \textsc{AdaGrad MD} have
predictions identical to \textsc{Scale-Free MD}. Hence the lower bound applies
to all four algorithms.
